Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-02 paper.tex
`Modify text to match new subgraph plots

\includegraphics[width=0.6\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
\\ % &
% \multicolumn{2}{c}{
\includegraphics[width=\textwidth]{figures/Pango_XAG_nxcld_tight_graph.pdf}
% }
\end{tabularx} \caption{\label{fig:pango-single-origin-graph} Single origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all XA samples
and their ancestral lineages are shown outwards to the nearest sampled viral
genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. (B) Details for the single-origin Pango X lineages in the
long ARG. (C) XAG origins in the long ARG.
} \end{figure}

In Figure~\ref{fig:pango-single-origin-graph}A we illustrate the case of XA. We construct a small subgraph of the long ARG, tracing routes from Nextclade-identified XA samples to the closest other sample strains in our dataset. Where a route passes upwards through a recombination node, we only show the continuing parental lineages to the most recent sample ancestors, rather than descending the ARG downward to cousins of the recombinant parents. Dotted lines show where this subgraph links to other nodes within the ARG. It is clear from the subgraph that all the XA samples trace to a single originating recombination node.

Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the
Pango recombinant lineages that trace back to a single recombination node in
the long ARG. [FIXME UPDATE THESE] We can see that XA, XAC, XAE, and XF form
monophyletic clades, in which all of the samples given that Pango designation
descendand from a single recombination node.

@@

[Discuss panel C where XAG shares a common recombinant origin with samples from
several other X lineages] All XAG samples share a single origin; however this
is also shared with all XAA samples, and includes some, but not all, of the
samples identified as XAB, XQ, and XU by Nextclade. Note that if the GISAID
designations are used, many of the samples marked here as XAB are reclassified
as BA.2, and XAG becomes fully monophyletic (supplementary figure
\ref{fig:pango_XAG_gisaid_graph}; this also gives identifying labels to the
sample nodes, which can be converted to EPI\_ISL indentifiers using
Supplementary table XXX )

\subsubsection{Multiple origin} The ARG inference has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. In this section we examine the Pango X lineage
samples that are inferred to have multiple recombinant origins (i.e., trace
back to different recombination nodes in the ARG), but still form monophyletic
clades % is this the right term? descending from these nodes .

@@

\begin{figure}
\begin{tabular}{cc}

\includegraphics[width=0.45\textwidth]{figures/Pango_XD_nxcld_tight_graph.pdf}

 &

\begin{tabular}{llll}
\toprule
Left parent & Right parent & \multicolumn{2}{l}{Breakpoint interval}\\
\midrule
BA.1.15 & BA.2  &9867  &10198\\
BA.1.15 & BA.2  &9867  &10198\\
BA.1.1  &BA.2  &10451  &11537\\
BA.1.1.5 & BA.2 &10448 & 11537\\
BA.1.15 & BA.2  &10448  &12880\\
BA.1.17 & BA.2  &10448 & 11537\\
BA.1  &BA.2.9  &10448 & 11537\\
BA.1  &BA.2  &12764  &13195\\
\bottomrule
\end{tabular}
% BA.3  &BA.1.17  &22674  &22674\\
% BA.1.1.15 Recombinant 10448 11537
% BA.1.1 Recombinant 10030 10447
% Recombinant BA.2 17411 19955
\end{tabular}
\caption{\label{fig:multiple_origins_graph}
A. Example subgraph for XD in the long ARG, which shows two origins for the
four Nextclade-labelled XD samples in the long ARG. B. details of inferred
breakpoints for different recombination events inferred for XE samples (170
samples and 12 inferred recombination nodes; some are very similar (Official:
BA.1* \& BA.2, 10448–11287)}
\end{figure}

In Figure~\ref{fig:multiple_origins_graph} we give the details of these
multiple pango lineages, along with an example lineage subgraph. In this XD
example the left hand subgraph (the origin of three XD samples) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL 8514045 (dated 2021-12-30). These
two samples differ by 35 mutations; considering the time between the samples,
it does seem possible that they represent true independent origins of XD.
Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the same subgraphs
with sample identifiers and coloured by GISAID Pango labels (however, no
samples are designated by GISAID as XD in the long ARG).

\subsubsection{Complex origins}
\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
\caption{\label{fig:complex_origins_graph}  DESCRIBE ME}
\end{figure}

As well as cases where a single recombination event explains several X lineages
(such as XAG above), many of the Pango X lineages involve more complex origins,
in which samples which have not been previously identified as recombinants also
descend from originating recombination events. In addition these complex
topologies often include extra recombination nodes within the topology that
descends from the main originating event. An example is XB in the long ARG,
plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
"hairball" of further recombination nodes whose descendants have not been
identified as recombinants by Nextclade (blue nodes, "pre-Alpha" VoC). However,
in this subgraph, there is only a single sample, identified as XB by Nextclade,
which is descended from these nested recombination events; moreover, this
sample is not identified as XB by GISAID (Supplementary
Figure~\ref{fig:pango_XB_gisaid_graph}). We suggest that investigating cases of
complex origins, and identifing which (if any) of the nested, further
recombination events may be artifactual, is an important area of future
research.

@@

\subsection{Parsimony increasing hueristics}
\label{sec:parsimony-heuristics}
Attaching trees built from the clusters of samples that copy from
a particular node (or path of nodes for recombinants, see the
next section) under the HMM is an inherently greedy strategy,
and can produce inferences that are clearly unparsimonious.
The final step in adding a daily batch of samples to the ARG
is then to perform some local updates that target specific
types of parsimony violation in the just-updated regions of the
ARG. There are currently two parsimony-increasing operations
applied, which we refer to as ``mutation collapsing'' and ``reversion
pushing''.

@@

Together these operations substantially increase the overall parsimony
of the inferred ARGs, most noticably in terms of the number of
reversion mutations inferred. However, they are both simple
greedy operations, just examining the local parts of the
ARG topology affected by newly added samples. Since the inferred
ARGs still contain a large number of reversion mutations which
are likely to be mostly artefactual, it is clear that there
is room for improvement, and that further parsimony increasing
hueristics would likely be of benefit.

>>>>

\includegraphics[width=\textwidth]{figures/Pango_XA_XAG_XD_nxcld_tight_graph.pdf}
% }
\end{tabularx} \caption{\label{fig:pango-simple-origin-graph} Simple origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. (B) Equivalent plot for the 17 XAG samples in the long ARG
(C) Equivalent plot for both origination events for the four XD samples in the long ARG.
} \end{figure}

In Figure~\ref{fig:pango-simple-origin-graph}A we illustrate the case of XA. We construct a small subgraph of the long ARG, tracing routes from Nextclade-identified XA samples to the closest other sample strains in our dataset. Where a route passes upwards through a recombination node, we only show the continuing parental lineages to the most recent sample ancestors, rather than descending the ARG downward to cousins of the recombinant parents. Dotted lines show where this subgraph links to other nodes within the ARG. It is clear from the subgraph that all the XA samples trace to a single originating recombination node.

Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the
Pango recombinant lineages that trace back to a single recombination node in
the long ARG. [FIXME UPDATE THESE] We can see that XA, XAC, XAE, and XF form
monophyletic clades, in which all of the samples given that Pango designation
descend from a single recombination node.

@@

[Discuss panel B where XAG shares a common recombinant origin with samples from
several other X lineages]

Figure~\ref{fig:pango-simple-origin-graph}B shows a case of another
Pango lineage, XAG, whose Nextclade-designated samples also trace to a
single originating recombination. This is a more intricate example, as the same origin
is also shared with all samples designated XAA, and some, but not all, of the
samples identified as XAB, XQ, and XU by Nextclade. Note, however, that if the GISAID
designations are used, many of the samples marked here as XAB are reclassified
as BA.2, and XAG becomes fully monophyletic (supplementary figure
\ref{fig:pango_XAG_gisaid_graph}, which also gives identifying labels to the
sample nodes, convertable to EPI\_ISL identifiers using
Supplementary table XXX )

\subsubsection{Multiple origins} The ARG inference has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. In this section we examine the Pango X lineage
samples that are inferred to have multiple recombinant origins (i.e., trace
back to different recombination nodes in the ARG), but still form monophyletic
clades % is this the right term? descending from these nodes .

@@

\begin{figure}

\begin{tabular}{llll}
\toprule
Left parent & Right parent & \multicolumn{2}{l}{Breakpoint interval}\\
\midrule
BA.1.15 & BA.2  &9867  &10198\\
BA.1.15 & BA.2  &9867  &10198\\
BA.1.1  &BA.2  &10451  &11537\\
BA.1.1.5 & BA.2 &10448 & 11537\\
BA.1.15 & BA.2  &10448  &12880\\
BA.1.17 & BA.2  &10448 & 11537\\
BA.1  &BA.2.9  &10448 & 11537\\
BA.1  &BA.2  &12764  &13195\\
\bottomrule
\end{tabular}
% BA.3  &BA.1.17  &22674  &22674\\
% BA.1.1.15 Recombinant 10448 11537
% BA.1.1 Recombinant 10030 10447
% Recombinant BA.2 17411 19955
\caption{\label{tab:multiple_origins_table}
details of inferred
breakpoints for different recombination events inferred for XE samples (170
samples and 12 inferred recombination nodes; some are very similar (Official:
BA.1* \& BA.2, 10448–11287)}
\end{figure}

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple example, consisting of
the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
(the origin of three XD samples) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL 8514045 (dated 2021-12-30). These
two samples differ by 35 mutations; considering this, and the time between the samples,
it does seem possible that they represent true independent origins of XD.
Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the same subgraphs
with sample identifiers and coloured by GISAID Pango labels (however, no
samples are designated by GISAID as XD in the long ARG).

Table \ref{tab:multiple_origins_table} summarises other Pango recombinants which
we infer to have multiple origins. [TODO: needs more detail]

\subsubsection{Complex origins}
\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
\caption{\label{fig:complex_origins_graph}  A Pango lineage subgraph showing
complex origins of the recombinant lineage XB. All XB samples in the long ARG
trace to a single recombination node (top centre), but three further recombinations
also descend from this node, with many samples descending from these nested
recombinations not being identified as XB, but instead assigned by Nextclade to
various pre-Alpha lineages.}
\end{figure}

As well as cases where single recombination events explain an X lineage,
such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, many of the
Pango X lineages involve more complex origins,
in which additional recombination events are nested inside the earliest shared
recombination. Frequently, the descendants of the original recombination
also include samples which have not been previously identified as recombinants.
An example is XB in the long ARG,
plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants have not been
identified as recombinants by Nextclade (blue nodes, ``pre-Alpha'' VoC). However,
in this subgraph, there is only a single sample, identified as XB by Nextclade,
which is descended from these nested recombination events; moreover, this
sample is not identified as XB by GISAID (Supplementary
Figure~\ref{fig:pango_XB_gisaid_graph}). We suggest that investigating cases of
complex origins, and identifying which (if any) of the nested, further
recombination events may be artifactual, is an important area of future
research.

@@

\subsection{Parsimony increasing heuristics}
\label{sec:parsimony-heuristics}
Attaching trees built from the clusters of samples that copy from
a particular node (or path of nodes for recombinants, see the
next section) under the HMM is an inherently greedy strategy,
and can produce inferences that are clearly unparsimonious.
The final step in adding a daily batch of samples to the ARG
is then to perform some local updates that target specific
types of parsimony violation in the just-updated regions of the
ARG. There are currently two parsimony-increasing operations
applied, which we refer to as ``mutation collapsing'' and ``reversion
pushing''.

@@

Together these operations substantially increase the overall parsimony
of the inferred ARGs, most noticably in terms of the number of
reversion mutations inferred. However, they are both simple
greedy operations, just examining the local parts of the
ARG topology affected by newly added samples. Since the inferred
ARGs still contain a large number of reversion mutations which
are likely to be mostly artefactual, it is clear that there
is room for improvement, and that further parsimony increasing
heuristics would likely be of benefit.
